---
title       : Pitch - Pitch
subtitle    : A Shiny PitchFX Visualization
author      : 
job         : 
framework   : io2012     # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## The Real MLB Strike Zone

- Question: does the real Major League Baseball STRIKE ZONE, as called by umpires, conform to the MLB Rulebook's definition?
  
  - If not, what are the true shape and dimensions of the strike zone?

- Major League Baseball Rulebook, Rule 2.00, The STRIKE ZONE   
<http://mlb.mlb.com/mlb/official_info/umpires/rules_interest.jsp>:

  - "The STRIKE ZONE is that area over home plate the upper limit of which is a horizontal line at the midpoint between the top of the shoulders and the top of the uniform pants, and the lower level is a line at the hollow beneath the kneecap. The Strike Zone shall be determined from the batter's stance as the batter is prepared to swing at a pitched ball."

---

## The Shiny App

The Shiny app has two main parts (switched via the radio button):

1. Plots of pitch densities -- the actual location of pitches thrown -- and colored by whether those pitches were typically called a ball or strike.

2. Plots of a General Additive Model which show the probability that a pitch thrown in a certain location will be called a strike.

- The data are from the publicly-available PitchFX database.

- Options
  - The drop down menu allows the user to analyze any of a selection of MLB pitchers.
  - You can toggle the rulebook-defined strike zone.
  - You can adjust the bounds of the x- and y-axes.

--- &twocol

## Pitch Density Plot

*** =left

- The dotted black line represents the MLB Rulebook definition of the strike zone.

- The orange contour lines represent increasing densities of called strikes. 

- The "dodger blue" contour lines indicate increasing densities of balls. 

- Each densitiy is emphasized by the background coloring of the plot.

*** =right
```{r density.plot, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

## load all pitch data
pitch_data <- read.csv("data/pitch_data2.csv", header=TRUE, stringsAsFactors=FALSE)

## guideline strike zone dimensions
sz.bot <- 1.6
sz.top <- 3.6
sz.left <- -0.95
sz.right <- 0.95
sz.dim <- data.frame(cbind(px = c(sz.left, sz.right, sz.right, sz.left, sz.left), 
                           pz = c(sz.bot, sz.bot, sz.top, sz.top, sz.bot)))

## density.plot function
density.plot <- function(name, zone, x.min, x.max, y.min, y.max) {
        
        data <- subset(pitch_data, pitch_data$pitcher_name == name)
        
        # split left and right-handed batters
        data.l <- data[data$stand == "Left-handed Batter", ]
        data.r <- data[data$stand == "Right-handed Batter", ]
        
        # function for plotting densities of Balls and Called Strikes
        density <- function(e) {
        
                cs <- e[e$des == "Called Strike", ]
                ball <- e[e$des == "Ball", ]
        
                ggplot(rbind(data.frame(cs, group="Strike"), data.frame(ball, group="Ball")), aes(x=px,y=pz)) +
                        stat_density2d(geom="tile", aes(fill = group, alpha=(..density..)*1.67), n=30, contour=FALSE) +
                        geom_path(data=sz.dim, aes(px,pz), linetype=5, alpha=zone) +
                        scale_fill_manual(values=c("Strike"="orange", "Ball"="dodgerblue")) +
                        scale_color_manual(values=c("Strike"="#FF4719", "Ball"="#3333CC")) +
                        stat_density2d(aes(alpha=(..level..)*1.67, colour=group), contour=TRUE) +
                        xlim(x.min, x.max) + ylim(y.min, y.max) +
                        theme_minimal() + 
                        theme(aspect.ratio=1, strip.text.x=element_text(size=12), legend.position="bottom", 
                                plot.title=element_text(size=16), axis.title=element_text(size=12), 
                                axis.title.x=element_text(vjust=0), axis.title.y=element_text(vjust=1)) + 
                                xlab("Horizontal Location") + ylab("Height") +
                        guides(fill=guide_legend(""), alpha="none", color="none")
        }

p1 <- density(data.l) + ggtitle("Left-handed Batters")
# p2 <- density(data.r) + ggtitle("Right-handed Batters")
# grid.arrange(p1, p2, ncol=2)
p1

}

density.plot("Jon Lester", 0.8, -2.5, 2.5, 0, 5)
```

--- &twocol

## GAM Model Plot

*** =left

- Again, the dotted black line represents the MLB Rulebook definition of the strike zone.

- Each color band represents a different probability that a pitch in a specific location will be called a strike.

- The boundary between the green and blue bands forms the shape of the modeled strike zone. At this boundary, there is a 50/50 chance of the pitch being called a strike or ball.

*** =right
```{r model.plot, echo=FALSE, message=FALSE, warning=FALSE}
library(gam)

model.plot <- function(name, zone, x.min, x.max, y.min, y.max) {
        
        # subset only called pitches
        data <- subset(pitch_data, pitch_data$pitcher_name == name)
        
        # split left and right-handed batters
        data.l <- data[data$stand == "Left-handed Batter", ]
        data.r <- data[data$stand == "Right-handed Batter", ]
        
        summary(data.l)
        
        # fit model, predict strikezone boundaries
        pred <- function(d) {
                
                aspect.ratio <- (max(d$px)-min(d$px))/(max(d$pz)-min(d$pz))
                d$des <- as.factor(d$des)
                
                # fit gam model
                fit <- gam(data=d, des ~ lo(px, span=.5*aspect.ratio, degree=1) + 
                                       lo(pz, span=.5, degree=1), family=binomial(link="logit"))
                
                # set up matrices for prediction
                myx.gam <- matrix(data=seq(from=-2, to=2, length=30), nrow=30, ncol=30)
                myz.gam <- t(matrix(data=seq(from=0,to=5, length=30), nrow=30, ncol=30))
                fitdata.gam <- data.frame(px=as.vector(myx.gam), pz=as.vector(myz.gam))
                
                # predict strikezone boundaries
                mypredict.gam <- predict(fit, fitdata.gam, type="response")
                mypredict.gam <- matrix(mypredict.gam, nrow=c(30,30))
                
                df <- data.frame(fitdata.gam, z=as.vector(mypredict.gam))
                df
        }
        
        df.l <- pred(data.l)
        df.r <- pred(data.r)        
        
        model <- function(df) {
                # model plot -- polygons from stat_contour
                ggplot(df, aes(x=px, y=pz)) +
                        stat_contour(geom="polygon", bins=8, aes(z=z, fill=..level..)) +
                        scale_fill_gradientn(colours = topo.colors(10), guide=guide_colorbar(title="P(Strike)", barwidth=12, nbin=4, draw.llim=TRUE, draw.ulim=TRUE)) +
                        geom_path(data=sz.dim, aes(px, pz), colour="black", alpha=zone, linetype=5) +
                        xlim(x.min, x.max) + ylim(y.min, y.max) +
                        theme_minimal() +
                        theme(aspect.ratio=1, strip.text.x=element_text(size=12), legend.position="bottom", 
                                plot.title=element_text(size=16), axis.title=element_text(size=12), 
                                axis.title.x=element_text(vjust=0), axis.title.y=element_text(vjust=1)) + 
                                xlab("Horizontal Location") + ylab("Height")
                }
        
        q1 <- model(df.l) + ggtitle("Left-handed Batters")
#         q2 <- model(df.r) + ggtitle("Right-handed Batters")
#         grid.arrange(q1, q2, ncol=2)
q1
}

model.plot("Jon Lester", 0.8, -2.5, 2.5, 0, 5)
```
